Notes and other text

### 2023-03-30 doc/manuscript/translocation.qmd
Initial comments on genomics analysis

<!--# Roland will address some of Bree's points here -->Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant genotypes [@knapp2016].
Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018].
Here, we report the results of (i) an analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd, (ii) a greatly expanded MYL frog reintroduction effort, and (iii) population viability modeling to assess the long-term persistence of reintroduced populations.
Collectively, the results show that MYL frogs generally meet the four conditions for reintroduction feasibility outlined above, and provide a novel example of how the reintroduction of resistant frogs can allow the large-scale recovery of this disease-impacted amphibian.
These results may have broad implications for the many species worldwide that are threatened with extinction by emerging diseases.
<!--# Bree comment: All of this was written before we knew the genomics had something cool to say. But now that the genomics feel a bit more elevated, I think this paragraph should be more bold. For example, we might want more here or in the Discussion to make it clear how the results of each of these sections are super important on their own. And that combining them is really amazing. I think the current tone is appropriately factual, but at some point we need to be louder about the novelty involved and the incredible amount of work that this represents.  This version feels like, we did a,b,c and so MYLF meet the conditions for feasibility (which makes it sound like we've only just gotten to the *starting* point of the study. I think we need to go further than saying they meet the four conditions of a specific conservation action. This is where beefing up the conclusions about evolutionary rescue seem appropriate. Also, genomic analyses typically come last in our papers as a dive into possible mechanistic explanations for observed patterns in the field. I think the order of the primary sections makes sense given the way you set up the argument about evolutionary rescue. That said, the order combined with the tone feels like it's contributing to the sense that the genomics are a box to be checked rather than contributing something exciting to the overall story. Now that Allie has added new analyses, I think we could make a slightly bolder set-up about how the parts fit together to make a surprisingly strong whole -->

@@

To identify putative genes under selection and potentially related to increased frog resistance to Bd, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
Specifically, we compared genomes of frogs sampled in four populations that have never experienced a Bd-caused epizootic ("naive") versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees (@fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
These "recovering" populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities ("load") in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
<!--# Modify text/remove text here/elsewhere to avoid repetition of this. --> Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.
To detect regions of the genome under selection, we compared exome sequences of frogs from naive and recovering populations using (1) a multivariate linear mixed model to evaluate associations between population type and individual variants, and (2) a splined window approach to evaluate larger genomic regions showing differentiation between population types <!--# Reverse the order of these two objectives? -->.
We then identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01) , or a threshold for the W-statistic for the splined window analysis (4 standard deviations from the mean).

@@

To visualize the genomic relationship of our populations we created a PCA using the glPCA function in the adegenet R package [@jombart2008].
Next, to detect regions of the genome under selection between naive and persistent populations, we used a multivariate linear mixed model as implemented in the program GEMMA [@zhou2014] <!--# Improve wording in this and subsequent paragraph to clarify that the two approaches were used to identify regions under selection. -->.
GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics for our association test.
For this analysis we used a stringent data filter to include only SNPs with \<5% missing data (missing for no more than 2 individuals).
GEMMA then calculated the likelihood ratio statistic for the set of 148,307 high quality SNPs across 127 contigs.
We then calculated two statistical thresholds for identifying outliers using a Bonferroni correction for a significance level of 0.05 (liberal set) and 0.01 (stringent set).
Results were visualized using a Manhattan plot and qqplot.

To detect regions of the genome under selection <!--# Improve wording -->, we conducted a splined window analysis for two statistics: *F~ST~* and differential nucleotide diversity ($\pi_{diff}$) between frogs from naive and recovering populations.
First, we calculated per-site F~ST~ between the naïve and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
Next, we calculated per-site $\pi$ separately for individuals from the naive and recovering populations individuals using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
We concatenated the values for F~ST~ and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
We then used the package WINGEN R package [@bishop2023] <!--# Correct citation? --> to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
This method calculates where non-overlapping window boundaries should occur by identifying inflection points in the fitted spline, therefore balancing false positive and false negative results that occur using other window-based methods [@beissinger2015] .
This method also calculates a W-statistic allowing for outlier identification.
We identified outliers as those with a W-statistic \> overall mean + 4 \* standard deviation for F~ST~ or overall mean +/- 4 \* standard deviation for $\pi_{diff}$.
Outlier regions were then identified as those that were outliers in both analyses, meaning that they showed 1) high differentiation between naive and recovering populations, and 2) differential patterns of nucleotide diversity in the same region.
We then extracted gene transcripts mapped within each region and retrieved annotation for that region using blast.

>>>>

<!--# Roland will address some of Bree's points here -->

Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant genotypes [@knapp2016].
Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018].
Here, we report the results of (i) an analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd, (ii) a greatly expanded MYL frog reintroduction effort, and (iii) population viability modeling to assess the long-term persistence of reintroduced populations.
Collectively, the results show that MYL frogs generally meet the four conditions for reintroduction feasibility outlined above, and provide a novel example of how the reintroduction of resistant frogs can allow the large-scale recovery of this disease-impacted amphibian.
These results may have broad implications for the many species worldwide that are threatened with extinction by emerging diseases.
<!--# Bree comment: All of this was written before we knew the genomics had something cool to say. But now that the genomics feel a bit more elevated, I think this paragraph should be more bold. For example, we might want more here or in the Discussion to make it clear how the results of each of these sections are super important on their own. And that combining them is really amazing. I think the current tone is appropriately factual, but at some point we need to be louder about the novelty involved and the incredible amount of work that this represents.  This version feels like, we did a,b,c and so MYLF meet the conditions for feasibility (which makes it sound like we've only just gotten to the *starting* point of the study. I think we need to go further than saying they meet the four conditions of a specific conservation action. This is where beefing up the conclusions about evolutionary rescue seem appropriate. Also, genomic analyses typically come last in our papers as a dive into possible mechanistic explanations for observed patterns in the field. I think the order of the primary sections makes sense given the way you set up the argument about evolutionary rescue. That said, the order combined with the tone feels like it's contributing to the sense that the genomics are a box to be checked rather than contributing something exciting to the overall story. Now that Allie has added new analyses, I think we could make a slightly bolder set-up about how the parts fit together to make a surprisingly strong whole -->

@@

To assess whether frog populations display an evolutionary response to Bd invasion, we used exome capture methods to compare genomes of frogs in populations with different histories of Bd exposure.
Our goal was to identify putative genes under selection and potentially related to increased frog resistance to Bd.
Specifically, we compared genomes of frogs sampled in four populations that have never experienced a Bd-caused epizootic ("naive") versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees (@fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
These "recovering" populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and Bd infection intensities ("load") in adults that are typically well below the level expected to cause frog mortality [@knapp2011, @vredenburg2010].
<!--# Modify text/remove text here/elsewhere to avoid repetition of this. --> Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.
To detect regions of the genome under selection, we compared exome sequences of frogs from naive and recovering populations using (1) a multivariate linear mixed model to evaluate associations between population type and individual variants, and (2) a splined window approach to evaluate larger genomic regions showing differentiation between population types <!--# Reverse the order of these two objectives? -->.
We then identified outlier variants using a Bonferroni-corrected significance level for single SNPs (liberal = 0.05, stringent = 0.01) , or a threshold for the W-statistic for the splined window analysis (4 standard deviations from the mean).

@@

To visualize the genomic relationship of our populations we created a PCA using the glPCA function in the adegenet R package [@jombart2008].
Next, to detect regions of the genome under selection between naive and persistent populations, we used a multivariate linear mixed model as implemented in the program GEMMA [@zhou2014] <!--# Improve wording in this and subsequent paragraph to clarify that the two approaches were used to identify regions under selection.  For example, for use genomics dummies it would be useful to see something like: "We used two related approaches to identify regions under selection. The first approach A, did X, Y, and Z but had C limitations.  The second approach B, did X, Y, and Z and D limitations."-->.
GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics for our association test.
For this analysis we used a stringent data filter to include only SNPs with \<5% missing data (missing for no more than 2 individuals).
GEMMA then calculated the likelihood ratio statistic <!--# What is the null with this test (or what are the two models being compared with the LRT)?  Probably obvious for genomics folks, but not totally clear to me.  I am guessing we are testing whether the `population` coefficient (naive vs. persistent) is significant for describing variability in a SNP? --> for the set of 148,307 high quality SNPs across 127 contigs.
We then calculated two statistical thresholds for identifying outliers using a Bonferroni correction for a significance level of 0.05 (liberal set) and 0.01 (stringent set).
<!--# Could we say something like: "Outliers indicated that a SNP was differently represented in naive vs. persistent populations" or whatever is the correct language." --> Results were visualized using a Manhattan plot and qqplot.

To detect regions of the genome under selection <!--# Improve wording.  Also, can we contrast this with the previous approach? What is this analysis doing that the other one didn't? -->, we conducted a splined window analysis for two statistics: *F~ST~* and differential nucleotide diversity ($\pi_{diff}$) between frogs from naive and recovering populations.
First, we calculated per-site F~ST~ between the naïve and recovering individuals for all bi-allelic SNPs in the 30 largest contigs (98% of all SNPs) using VCFtools [@danecek2011].
Next, we calculated per-site $\pi$ separately for individuals from the naive and recovering populations individuals using VCFtools, then calculated $\pi_{diff}$ for each population ($\pi_{diff} = \pi_{naive} - \pi_{recovering}$).
We concatenated the values for F~ST~ and $\pi_{diff}$ in order of size-sorted chromosome number and adjusted the SNP position based on the relative position in the genome.
We then used the package WINGEN R package [@bishop2023] <!--# Correct citation? --> to conduct a splined discrete window analysis for *F~ST~* and $\pi_{diff}$.
This method calculates where non-overlapping window boundaries should occur by identifying inflection points in the fitted spline, therefore balancing false positive and false negative results that occur using other window-based methods [@beissinger2015] .
This method also calculates a W-statistic allowing for outlier identification.
We identified outliers as those with a W-statistic \> overall mean + 4 \* standard deviation for F~ST~ or overall mean +/- 4 \* standard deviation for $\pi_{diff}$.
Outlier regions were then identified as those that were outliers in both analyses, meaning that they showed 1) high differentiation between naive and recovering populations, and 2) differential patterns of nucleotide diversity in the same region.
We then extracted gene transcripts mapped within each region and retrieved annotation for that region using blast.
